---
title: "Getting started with simOutrank"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with simOutrank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

`simOutrank` clusters the traces of an event log by an *outranking* model of
similarity. Instead of collapsing every perspective into a single distance, it
scores each pair of traces on several problem-specific criteria -- activities,
transitions, duration, case attributes -- and aggregates them with
ELECTRE-III-style concordance and discordance. The result is a credibility
matrix `S` that a spectral or hierarchical algorithm then partitions.

This vignette walks through the whole pipeline on the bundled
`illustrative_log`, a 25-case customer-service log from Delias et al. (2023).

```{r setup}
library(simOutrank)
head(illustrative_log)
```

## 1. Traces

`as_traces()` turns an event log into a `traces` object: the time-ordered,
integer-encoded activity sequence of every case, plus the case-level attribute
table.

```{r traces}
traces <- as_traces(illustrative_log,
                    case_id = "case_id", activity = "activity",
                    timestamp = "timestamp")
traces
summary(traces)[1:5, ]
```

## 2. Criteria

Criteria are built with the `crit_*` helpers. Each carries a direction
(similarity or dissimilarity), a weight, and ELECTRE-III thresholds
(indifference, similarity, and an optional veto). Here we mirror the four
criteria of the source paper: what activities occur, in what order, and two
case attributes.

```{r criteria}
criteria <- list(
  crit_activity_profile(weight = 0.2, indifference = 0.7, similarity = 0.8,
                        veto = 0.4),
  crit_edit_distance(weight = 0.2, similarity = 2, indifference = 3, veto = 6),
  crit_nominal("status", weight = 0.3),
  crit_nominal("satisfaction", weight = 0.3)
)
criteria[[1]]
```

Thresholds may be fixed numbers or quantiles of the criterion's own value
distribution via `q()`; the [parameter-tuning
vignette](parameter-tuning.html) explores that choice.

## 3. Similarity

`outrank_similarity()` resolves any quantile thresholds, normalises the
weights, and aggregates the per-criterion indices into `S`.

```{r similarity}
sim <- outrank_similarity(traces, criteria)
sim
round(sim$S[1:5, 1:5], 2)
```

## 4. How many clusters?

`eigengap()` plots the smallest eigenvalues of the normalized Laplacian; a gap
after the k-th value suggests k groups.

```{r eigengap}
eigengap(sim, k_max = 8)
```

## 5. Cluster

The paper targets four natural groups (tier x satisfaction). Spectral
clustering uses a seed for reproducible k-means.

```{r cluster}
clust <- cluster_traces(sim, k = 4, seed = 42)
clust
split(names(clust$memberships), clust$memberships)
```

## 6. Use the result

Join memberships back onto the event log, inspect a cluster's attribute
profile, and compute validity indices.

```{r use}
augmented <- augment_log(clust, illustrative_log)
head(augmented)

plot(clust, type = "profile", attribute = "status")

if (requireNamespace("clValid", quietly = TRUE)) {
  validate_clusters(clust)
}
```

## Where next

* [Reproducing the illustrative example](reproducing-paper2.html) -- the same
  pipeline framed as a published-result reproduction.
* [Robustness: constraints and trimming](robustness.html) -- inject domain
  knowledge and remove outliers.
* [Parameter tuning](parameter-tuning.html) -- weights, thresholds and k.

## Reference

Delias, P., Doumpos, M., Grigoroudis, E. and Matsatsinis, N. (2023). Improving
the non-compensatory trace-clustering decision process. *International
Transactions in Operational Research*, 30(3), 1387–1406.
doi:10.1111/itor.13062
```
