---
title: "Reproducing the emergency-room study (foundations paper)"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
library(simOutrank)
set.seed(2021)
```

> **Data note.** The foundations paper (Delias et al., 2017, *A non-compensatory
> approach for trace clustering*) uses a real emergency-room (ER) log that is
> not redistributable. This article reproduces the *modelling recipe* on a
> synthetic ER-like log with the same structure, and shows exactly how to swap
> in the real data. It is a pkgdown article, not a CRAN vignette, so it may use
> heavier data than the built vignettes.

## A synthetic ER log

The study clusters emergency-room patient traces on multiple perspectives:
which activities occur, their order, the triage colour (ordinal), the case
type and shift (nominal), and the duration. We synthesise three archetypes --
a fast track, a moderate path, and a critical path -- with noise.

```{r synth}
archetypes <- list(
  fast     = c("Arrival", "Triage", "Exam", "Discharge"),
  moderate = c("Arrival", "Triage", "Exam", "BloodTest", "BloodResult",
               "Decision", "Discharge"),
  critical = c("Arrival", "Triage", "Exam", "XRay", "XRayResult", "BloodTest",
               "BloodResult", "RoomEntrance", "RoomExit", "Discharge")
)
triage_of <- c(fast = "green", moderate = "yellow", critical = "red")

make_case <- function(id, kind) {
  acts <- archetypes[[kind]]
  if (runif(1) < 0.2 && length(acts) > 4)         # a little structural noise
    acts <- acts[-sample(seq(3, length(acts) - 1), 1)]
  n <- length(acts)
  step <- switch(kind, fast = 15, moderate = 40, critical = 90)
  data.frame(
    case_id   = id,
    activity  = acts,
    timestamp = as.POSIXct("2017-01-01", tz = "UTC") +
                cumsum(c(0, round(rexp(n - 1, 1 / step)))) * 60,
    triage    = unname(triage_of[kind]),
    type      = sample(c("medical", "surgical"), 1),
    shift     = sample(c("day", "night"), 1),
    stringsAsFactors = FALSE
  )
}

kinds <- rep(c("fast", "moderate", "critical"), times = c(18, 15, 12))
er_log <- do.call(rbind, Map(make_case, sprintf("P%02d", seq_along(kinds)),
                             kinds))
traces <- as_traces(er_log, "case_id", "activity", "timestamp")
traces
```

## Criteria across perspectives

The `crit_*` helpers cover every scale the study uses. Triage is ordinal
(green < yellow < red); type and shift are nominal; duration is quantitative.

```{r criteria}
criteria <- list(
  crit_activity_profile(weight = 21),
  crit_edit_distance(weight = 23, similarity = 2, indifference = 4, veto = 8),
  crit_ordinal("triage", levels = c("green", "yellow", "red"), weight = 16,
               similarity = 0, indifference = 1.5, veto = 2.5),
  crit_nominal("type", weight = 3),
  crit_nominal("shift", weight = 16),
  crit_duration("mins", weight = 12, similarity = q(0.2), indifference = q(0.6))
)
sim <- outrank_similarity(traces, criteria)
sim
```

Composite criteria from the paper (for example *Flow = Levenshtein - 14 x
transition similarity*) are expressed with `crit_custom()` and a user function
of `traces`; here we keep the perspectives separate for clarity.

## Three clusters

The paper reports a three-cluster solution. Spectral clustering recovers the
archetypes:

```{r cluster}
clust <- cluster_traces(sim, k = 3, seed = 100)
table(cluster = clust$memberships,
      triage  = er_log$triage[match(names(clust$memberships), er_log$case_id)])
```

Each cluster is dominated by one triage colour -- the fast/moderate/critical
structure the synthetic log was built from, recovered without ever telling the
algorithm the triage.

## Swapping in the real ER log

With the real CSV, only the ingestion changes; the criteria and clustering are
identical.

```{r real, eval = FALSE}
raw <- read.csv("Filter1_MergedPatientsLog.csv")
traces <- as_traces(raw,
                    case_id   = "Case.ID",
                    activity  = "Activity",
                    timestamp = "Complete.Timestamp")
# ...then the same `criteria` and `cluster_traces(sim, k = 3)` as above.
```

## Reference

Delias, P., Doumpos, M., Grigoroudis, E., Manthou, V. and Matsatsinis, N.
(2017). Supporting healthcare management decisions via robust clustering of
event logs. *Knowledge-Based Systems*.
```
